---
title: "TP1"
output: html_notebook
author: "Anaïs Artaud, Silvana Licaj, Hiroyoshi Yamasaki"
---

```{r}
library(ggplot2)
library(poisson)
library(STAR)
library(rjags)
library(R2jags)
```

# Part I: Generating random variables
- Generate a vector x of length 200 of equispaced values from 0 to 3 using the function ‘seq’. Using the function ‘dexp’, create a plot of the probability density function of an exponential distribution with parameter 5.

```{r}
x <- seq(0, 3, length=200)
y <- dexp(x, rate=5)
df <- data.frame(x=x, y=y)
p <- ggplot(data=df, aes(x=x, y=y)) + geom_line(col="blue") 
p + ggtitle("pdf of Exponential distribution") + xlab("value x") + ylab("density")
```

Draw a sample of 1000 values from an exponential distribution with parameter 5 using the function ‘rexp’. Draw a histogram using the function ‘hist’.

```{r}
samples <- rexp(n=1000, rate=5)
df <- data.frame(x=samples)
p <- ggplot(data=df, aes(x=x)) + geom_histogram() + theme(plot.title = element_text(hjust = 0.5))
p + ggtitle("Histogram of exponential distribution") + xlab("value x") + ylab("Frequency")
```

Create a vector of integers from 0 to 10. Use the function ‘barplot’ to represent the probability mass function of a Poisson distribution with parameter 3. Generate 1000 random draws from this distribution. Use the functions ‘table’ and ‘barplot’ to give the probability mass function of these generated values.

```{r}
x <- 0:10
y <- dpois(x, lambda=3)

df = data.frame(x=x, y=y)
p <- ggplot(data = df, aes(x=x, y=y)) + geom_bar(stat="identity") + theme(plot.title = element_text(hjust = 0.5))
p + ggtitle("pdf of Poisson distribution") + xlab("value x") + ylab("density")
```
Run the CODE chunk 1 below. Explain what it does and what principles it is illustrating.

```{r}
# Gamma distribution
par(cex.lab=2)       # magnification on x and y labels
par(cex.axis=1.75)   # magnification on axis annotations
par(las=1)           # style of axis label, horizontal
par(mfrow=c(1,2))    # 1-row, 2-cols
par(mar = c(4, 4, 1, 1))  # margin size

# Sample
small_sample = 10
large_sample = 200
smallSample = rgamma(small_sample, 2)
largeSample = rgamma(large_sample, 2)

# Small sample #################################################################

# Format data
xvals = sort(smallSample) # sort
yvals = (0:small_sample)/small_sample         # vector of 10 with increment .1

# Create a stepfunction
fhat = stepfun(xvals, yvals, right = TRUE)

# Plot the step function
plot(fhat, xlab = "", ylab = "", main = "", pch = 16, cex = 1)

# Plot the cdf
xseq = seq(-1, 12, length = 300) # input values x
lines(xseq, pgamma(xseq, 2), lty = 5, lwd = 2, col = 4)

# Large sample  ################################################################

# Format data
xvals = sort(largeSample)
yvals = (0:large_sample)/large_sample

# Create a stepfunction
fhat = stepfun(xvals, yvals, right = TRUE)

# Plot the step function
plot(fhat, xlab = "", ylab = "", main = "", pch = 16, cex = 0.7)

# Plot the cdf
xseq = seq(-1, 12, length = 300)
lines(xseq, pgamma(xseq, 2), lty = 5, lwd = 2, col = 4)
```

Adapt the CODE chunk 1 by replacing the Gamma distribution with a Gaussian distribution with
mean 0 and standard deviation 1 and the consider sample sizes of 30 and 150

```{r}
# Gaussian distribution
par(cex.lab=2)       # magnification on x and y labels
par(cex.axis=1.75)   # magnification on axis annotations
par(las=1)           # style of axis label, horizontal
par(mfrow=c(1,2))    # 1-row, 2-cols
par(mar = c(4, 4, 1, 1))  # margin size

# Sample
small_sample = 30
large_sample = 150
smallSample = rnorm(small_sample)
largeSample = rnorm(large_sample)

# Small sample #################################################################

# Format data
xvals = sort(smallSample) # sort
yvals = (0:small_sample)/small_sample         # vector of 10 with increment .1

# Create a stepfunction
fhat = stepfun(xvals, yvals, right = TRUE)

# Plot the step function
plot(fhat, xlab = "", ylab = "", main = "", pch = 16, cex = 1)

# Plot the density
xseq = seq(-10, 10, length = 300) # input values x
lines(xseq, pnorm(xseq), lty = 5, lwd = 2, col = 4)

# Large sample  ################################################################

# Format data
xvals = sort(largeSample)
yvals = (0:large_sample)/large_sample

# Create a stepfunction
fhat = stepfun(xvals, yvals, right = TRUE)

# Plot the step function
plot(fhat, xlab = "", ylab = "", main = "", pch = 16, cex = 0.7)

# Plot the density
xseq = seq(-10, 10, length = 300)
lines(xseq, pnorm(xseq), lty = 5, lwd = 2, col = 4)
```

# Part II: simulating a Poisson process

## Generating a Poisson Process

```{r}
generate_poisson <- function(n, lambda)
{
  out <- numeric(n+1) # starts with 0
  out[1] <- 0         # initialize
  
  for (i in 2:(n+1))
  {
    sample <- runif(1)
    xi <- -log(1 - sample) / lambda  # F^-1(u)
    out[i:(n+1)] <- out[i:(n+1)] + rep(xi, n-i+2)
  }
  return (out)
}

```

```{r}

n = 100
lambda = 5

x1 <- generate_poisson(n, lambda)
x2 <- hpp.sim(lambda, n)

df <- data.frame(c(x1, x2), c(1:(n+1), 1:(n+1)), c(rep("ours", n+1), rep("hpp", n+1)))
names(df) <- c("t", "count", "type")

p <- ggplot() + geom_step(data=df, aes(x=t, y=count, color=type)) + theme(plot.title = element_text(hjust = 0.5))
p + ggtitle("Comparison of 2 Poisson Process functions") + xlab("Time t") + ylab("Number of events")
```


```{r}

n = 100
lambda = 5
num_trials = 10

x <- numeric(0)
c <- numeric(0)
t <- numeric(0)

for (i in 1:num_trials)
{
  x <- c(x, generate_poisson(n, lambda))
  x <- c(x, hpp.sim(lambda, n))
  
  c <- c(c, 1:(n+1))
  c <- c(c, 1:(n+1))
  
  t <- c(t, rep("ours", n+1))
  t <- c(t, rep("hpp", n+1))
}


df <- data.frame(x, c, t)
names(df) <- c("t", "count", "type")

p <- ggplot() + geom_point(data=df, aes(x=t, y=count, color=type, alpha=.1)) + theme(plot.title = element_text(hjust = 0.5))
p + ggtitle("Comparison of 2 Poisson Process functions") + xlab("Time t") + ylab("Number of events")
```


# Part III: real data analysis

## Building Peri-Stimulus Time Histograms

Create a vector ‘N1_S1’ of spike event times of the first trial of the neuron 1.

```{r}
data("CAL1V")
N1_S1 <- CAL1V$`neuron 1`$`stim. 1`
attributes(N1_S1) <- NULL
N1_S1
```

Check the number of events observed at each trial (i.e., compute the length of sub-list)

```{r}
lapply(CAL1V$`neuron 1`, FUN=length)
```

Compute the range of values taken by the time of events
```{r}
lapply(CAL1V$`neuron 1`, FUN=range)
```

Create a Peri-Stimulus Time Histogram. The idea is to aggregate (average) data over trials into bins of predetermined width (for example 0.5s), normalize the unit spike to the second, Hint: use the function ‘hist’.
```{r}
# set the disjoint interval from the hist function
breaks = seq(0, 11, by=0.5) # number of spikes per half seconds!

# create a matrix of counts within each half second intervals
neuron1_mat_counts = matrix(nrow = 20, ncol = length(breaks)-1) 
for (i in 1:length(Neuron_1))
{
  neuron1_mat_counts[i, ] =  hist(Neuron_1[[i]],breaks = breaks, plot = FALSE)$counts
}

plot(seq(0.1, 10.9, by=0.5), 2*apply(neuron1_mat_counts, 2, mean), type='l', main = "")
```

## Beyond (homogeneous) Poisson Processes

Plot the cumulative count of spike events (for neuron 1 trial 1) as a function of the event times and overplot with the cumulative plot of your Poisson process with parameter λ. Choose λ as the average number of events per seconds over the 20 replicates.


```{r}

lambda <- mean(unlist(CAL1V$`neuron 1`))
n <- length(N1_S1)
x <- generate_poisson(n-1, lambda=lambda)

df <- data.frame(c(N1_S1, x), c(1:n,1:n), c(rep("neuron", n), rep("poisson", n)))
names(df) <- c("t", "count", "type")
ggplot(data=df, aes(x=t, y=count, color=type)) + geom_step()
```

## Bayesian modeling
Using R, create 2 vectors of counts giving the number of spikes for each trial in time intervals $I_1 = (3, 4]$ and $I_2 = (5, 6]$. Denote as $y_{i,j$} the number of spikes for the $i$-th trial and $j$-th time interval.
```{r}
N1 <- CAL1V$`neuron 1`

count <- function(spikes, min, max)
{
  return (length(spikes[spikes > min & spikes <= max]))
}

y1 <- lapply(N1, 3, 4, FUN=count)
attributes(y1) <- NULL

y2 <- lapply(N1, 5, 6, FUN=count)
attributes(y2) <- NULL
```

You decide to use a Bayesian approach and propose the following model
$$
y_{i,j}|  \sim  Poi(\theta_j), j \in \{1, 2\} \\
\theta_j \sim Gamma(0.1, 0.1)
$$

Discuss this modeling choice.
Here we have a process of generalization of the Poisson law : the law is said to be inhomogeneous, that is to say that the parameter of the law (lambda) will vary and will no longer be fixed at a single value. Also, our lambda here follows a distribution of gamma law which justifies a combination of poisson - gamma law. 


Using the Rjags code chunk below, compute the posterior distribution of the differences of the Poisson parameters.

```{r}
# create the vector of counts
# Prepare Jags function input
model_code = '
model
{
  # likelihood
  for (i in 1:length(y1))
  {
  
    y1[i] ~ dpois(theta1)
    y2[i] ~ dpois(theta2)
      
  }
  # Prior
  theta1 ~ dgamma(0.1, 0.1)
  theta2 ~ dgamma(0.1, 0.1)
  delta <- theta1 - theta2
} '
model_data = list(y1 = y1, y2 = y2,  N = 20)
model_parameters =  c("theta1","theta2","delta")
# run Jags
model_run = jags(data = model_data,
                 parameters.to.save = model_parameters,
                 model.file=textConnection(model_code),
                 n.chains=4,
                 n.iter=10000,
                 n.burnin=200,
                 n.thin=2)
# print the results
print(model_run)

# plot the posterior distribution of the parameter of interest
hist(model_run$BUGSoutput$sims.array[,1,"theta1"], main="", probability = TRUE, xlab="delta")
hist(model_run$BUGSoutput$sims.array[,1,"theta2"], main="", probability = TRUE, xlab="delta")
```

Give the histogram of the differences

```{r}
hist(model_run$BUGSoutput$sims.array[,1,"delta"], main="", probability = TRUE, xlab="delta")
```

From the Jags outputs what are your conclusions?
This result indicates that the assumption about the homogeneity was wrong and it explains why it diverges from the Poisson process.

